---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# KMeansStability

<!-- badges: start -->
<!-- badges: end -->

The goal of KMeansStability is to implement the functions necessary to conduct a small simulation study comparing the efficacy of a set of stability-based methods for specifying $K$ for $K$-mean clustering. In brief, stability-based specification methods are based on the principle that a correct specification of $K$ should induce the most reproducible clusters: if one were to repeatedly sample from the population, the clusterings tuned to the correct value of $K$ would be more similar compared to clusterings with inappropriate values of $K$. To leverage this intuition into a method, one emulates repeated samplings through resampling from a given dataset and defines a  distance metric over the resampled clusterings. The value of $K$ which yields the smallest average distance between the clusters is deemed the most stable tuning of $K$. The choice of distance metric is essential to defining the most stable value of $K$. KMeansStability implements two common stability metrics, Minimum Matching Distance and Comembership Distance, along with three novel distances based on the Wasserstein distance between different components of the clusterings, within a resampling framework which accomodates either cross-validation or the bootstrap for resampling.



## Installation

You can install the development version of KMeansStability from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wolfgangkbri/KMeansStability")
```

## Example

The main function in `KMeansStability` is `stability_analysis`. This function evaluates a set of five different stability metrics across a range of Kmeans clusterings. 

```{r example}
library(KMeansStability)
## basic example code
set.seed(3)
df <- toy_datagen()
results <- stability_analysis(df, Kmax = 6, method = "CV", nreps = 4)
results$mean_stabilities
```

With this dataset, we have that all five of the stability metrics agree that setting $K=3$ induces the most stable clustering, correctly identifying the number of components of the mixture model forming the dataset.

